\scuba: iterative map-making with the Sub-Millimetre User Reduction Facility

### 2012-09-03 smurf_paper.tex
Remove step fixing appendix

Detecting and correcting such a wide variety of steps reliably has proved
to be a challenge. The current step-correction algorithm is described in
detail in Appendix~\ref{appendix:steps}. In outline, the following stages
are involved in detecting steps in a single bolometer time stream:

@@

%------------------------------------------------------------------------------
\appendix
\section[]{Step correction in detail}
\label{appendix:steps}

An outline of the step correction algorithm is given in
Section~\ref{sec:steps}. This appendix contains a more detailed
description.

Detecting and correcting steps in a single bolometer time stream involves:

\item locating each candidate step;

\item determining the widths of the candidate steps;

\item determining the height of the step (steps can vary enormously in
height -- for instance the step in the left-hand panel of
Fig.~\ref{fig:steps1} is roughly 3\,pW in height, whereas the step at
sample 5100 in the right-hand panel is only 0.01\,pW in height -- the
height of a step is taken as the difference in the local median value
just before and just after the step);

\item correcting all subsequent data values by the step height;

\item once all steps have been corrected, a global offset is added to the
corrected bolometer time stream to ensure that the mean value before and
after step correction is unchanged.

\subsection{Detecting candidate steps}
The first stage in the algorithm is to search the bolometer time stream for
contiguous blocks of samples that have unusually large mean gradients
compared to the neighbouring samples. Each such block of samples is
considered to be a candidate step. The starting and ending time of each
such block is noted. Once all candidate steps have been found for a
bolometer, the algorithm proceeds to the next stage. Candidate steps are
found as follows.

\item Smooth the bolometer time stream using a median filter of width
50 samples. This reduces the effects of the noise in the data. Using a
median filter rather than a mean filter helps to preserve the sharp edges
at each step, and is robust against spikes, etc.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{steps2.pdf}
\caption{A section of the bolometer signal from the right-hand panel
of Fig.~\ref{fig:steps1} with the median smoothed data overlaid in
red.}
\label{fig:steps2}
\end{figure}

The red curve in Fig.~\ref{fig:steps2} shows the median smoothed data for
the right hand bolometer from Fig.~\ref{fig:steps1}.

\item At each time sample, find the difference between the two adjacent
values in the median filtered bolometer data. This is proportional to the
gradient of the median filtered bolometer data. Fig.~\ref{fig:steps3}
shows these differences for the sample bolometer shown in previous
figures. The high gradients at the three step are clearly visible at
indices 5040, 5100 and 5280.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{steps3.pdf}
\caption{Differences between adjacent values of the median smoothed
data shown in Fig.~\ref{fig:steps2}.}
\label{fig:steps3}
\end{figure}

\item Steps are located by looking for difference values that exceed 25
times the RMS value of the nearby neighbouring differences. However, the
local RMS of the differences includes the gradient of the underlying
noise-free time series. So if there is a strong upward or downward trend
in the time series data, the RMS of the difference values will be higher,
potentially resulting in some steps being missed. For instance, at sample
5000 the overall trend in the time series is downwards, resulting in a
negative bias to the difference values.\footnote{This bias is not visible in
Fig.~\ref{fig:steps3} because of the large vertical range, but can be seen in
Fig.~\ref{fig:steps4} which has a smaller vertical scale.}

Therefore to enhance the reliability of the step detection, such biases
are removed before locating the steps. To do this, the differences are
smoothed using a mean filter, and the smoothed differences are removed.
The remaining residual difference values represent the gradient due to
steps, spikes, noise, etc. It is important that the smoothed differences
are not polluted by the large values caused by spikes and steps. So
before smoothing, differences of more than three times the RMS difference
(taken over the whole time stream) are identified and flagged. The RMS is
then re-calculated omitting the flagged values and a further two such
rejection iterations are performed. This process flags high gradient values
caused by spikes, jumps, point sources, etc. The difference values close
to a step are often not typical of those in the wider neighbourhood, so
each block of contiguous flagged difference values is doubled in width.
These flagged values are shown as black in Fig.~\ref{fig:steps4}, with
the remaining unflagged values shown in red.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{steps4.pdf}
\caption{The red curve shows the data from Fig.~\ref{fig:steps3} plotted with smaller upper
and lower limits. The blue curve shows the results of smoothing the red
curve. Points omitted from the red curve are shown in black.}
\label{fig:steps4}
\end{figure}

\item The differences for the remaining samples are smoothed
using a mean filter of width 50 samples. The smoothing process fills in small
holes in the array (up to 40 samples), but will not fill in larger holes.
Such larger gaps are filled in later by linear interpolation. The
resulting smoothed differences (shown in blue in Fig.~\ref{fig:steps4})
provide an estimate of the local gradient of the underlying noise-free
median-filtered data stream.

\item The smoothed differences are subtracted from the total differences
to get the residual differences caused by noise, spikes, steps, etc.
These are show in black in Fig.~\ref{fig:steps5}.

\item The local RMS of the residual differences at each time sample is
found by smoothing the squared residual differences using a median filter
of width 200 samples, and then taking the square root of the smoothed values.
These are show in green in Fig.~\ref{fig:steps5}.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{steps5.pdf}
\caption{The black curve shows the residuals between the red and blue curves in
Fig.~\ref{fig:steps4}. Missing data is filled in by a combination of
smoothing and linear interpolation. The green curve is the local RMS of the data
over a box width of 200 samples.}
\label{fig:steps5}
\end{figure}

\item Blocks of contiguous samples for which the residual difference
exceeds 25 times the local RMS of the residual differences are
found. Any two such blocks that are separated by fewer than 100
samples are merged into a single block. Blocks are only kept if they
pass all of the following tests:

\item the width of the block must be less than 200 samples;

\item the total of all the \snr\ values (the absolute residual difference
divided by the local RMS of the residual differences) in the block must be
at least 25;

\item the total of all \snr\ values in the block must exceed half the value of
the largest single \snr\ value in the block.

The second and third tests are intended to prevent spikes being treated as
two very close steps, since such steps will have \snr\ values of opposite
signs and will thus cancel out.

The data in Fig.~\ref{fig:steps5} results in two blocks being defined,
one between samples 5037 and 5102, and another between samples 5286
and 5296. Each of these blocks define the start and end of a candidate
step.  The first block includes two visible steps, but they have been
merged into a single candidate step due to their small separation
(less than 100 samples).

\subsection{Measuring and confirming candidate steps}

%------------------------------------------------------------------------------

>>>>

Detecting and correcting such a wide variety of steps reliably has proved
to be a challenge. In outline, the following stages
are involved in detecting steps in a single bolometer time stream:

@@
